---
title: "An Interactive Map of Bigfoot Sightings"
author: "Claire McFarland"
format: html
editor: visual
embed-sesources: TRUE
---

## **Abstract**

Bigfoot is one of North America's most famous cryptids. First spotted in 1958, sasquatch sightings since have numbered in the thousands, with varying levels of scientific credibility. A foundation, the Bigfoot Field Researchers' Organization (BFRO) was founded in 1995 to head investigations into this hairy curiosity. Thanks to the Github blog TidyTuesday scraping data from the BFRO's website, we could perform several analyses. First was a map of pins where each encounter occurred. Second, a density map of each U.S. state shaded by the number of sightings. Thirdly, we generated a line plot displaying the cumulative number of sightings in the top five states with the most sightings. Interestingly, it appears that Washington has more or less always had the lead in sightings, at least among the other leaders. Oregon and California, too, have always had a high number of sightings, but since 2010, they both appear to have abruptly plateaued. Lastly, I performed a linear regression to observe if the climate variables provided by the BFRO significantly impact the credibility of an encounter. The notable observation from this visualization was that encounters with clear weather, of course, have the highest credibility, but they are somewhat comparable to encounters that occurred on snowy days, with rainy days (across all intensities) appearing to have the least credibility. This may be because heavy snow has the potential to leave lasting footprints, which the BFRO might consider stronger evidence.

Here I read in the data from TidyTuesday:

```{r}
#| message: false
#| warning: false

# install.packages("tidytuesdayR")

tuesdata <- tidytuesdayR::tt_load('2022-09-13')
tuesdata <- tidytuesdayR::tt_load(2022, week = 37)

bigfoot <- tuesdata$bigfoot

# Or read in the data manually

bigfoot <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2022/2022-09-13/bigfoot.csv')

```

```{r}
#| message: false
#| warning: false
# install.packages("leaflet")
# install.packages("tigris")
# install.packages("USABoundaries")
# install.packages("ggeffects")
library(leaflet)
library(readr)
library(dplyr)
library(lubridate)
library(tigris)
library(sf)
library(ggplot2)
library(broom)
library(ggeffects)
library(plotly)
```

## **Introduction**

*Do you believe in Bigfoot?* Possibly the most famous North American cryptid, Bigfoot sightings have become so ubiquitous that an entire organization of researchers was founded to investigate the credibility of these claims, and even venture into describing the ecology of their alleged creature. This organization is known as the BFRO (Bigfoot Field Researchers Organization).

Every reported Bigfoot sighting is stored in the BFRO's website, bfro.net. Here, BFRO's scholars record the witness's story on the encounter, collect information on variables related to climate and location, and classify the encounter into Class A, Class B, or Class C. Class A encounters are the most definitive, ruling out animals, lying/impaired witnesses, or otherwise "misidentifications". Class B encounters are less credible, with the potential for misidentifications or unclear evidence. And Class C encounters may be second or third-hand accounts, with little or untraceable evidence.

On September 13th, 2022, TidyTuesday provided a dataset scraped from the BFRO's catalogue of Bigfoot encounters to their weekly blog. This dataset contains many of the aforementioned climate and location variables as well as their credibility class and- *very* helpfully- the approximate latitude and longitude of the encounter.

With over 5,000 entries in this dataset, and 28 of these variables, the opportunities for visualizing Bigfoot sightings are endless. But my prevailing question is: Where and When does this tall, furry, cryptid like to show his face? During what kinds of conditions is he likely to be caught on a stroll? And where exactly do he and his fellow sasquatches live?

## **Data Tidying**

For the map with the stories of sightings:

```{r}
bigfoot_stories <- bigfoot %>%
  filter(!is.na(observed)) %>%
  filter(!is.na(location_details))
```

\*note: 789 observations were removed at this stage for having missing observation text and location detail.

## **Visualizations**

The interactive map below was made using leaflet. It simply displays every bigfoot sighting that has longitude and latitude information as well as the "story" behind the encounter and more descriptive location information.

```{r}
content <- bigfoot_stories %>%
  mutate(popup = paste0('<a href =', bigfoot_stories$location_details, '>',
                        bigfoot_stories$observed, '</a>'))
```

```{r}
bigfoot_map <- leaflet(bigfoot_stories) |>
  setView(lng = -98.583, lat = 39.833, zoom = 1) |>
  addTiles() |>
  addProviderTiles(providers$Esri.WorldPhysical) |>
  addMarkers(lng = bigfoot_stories$longitude, lat = bigfoot_stories$latitude,
             clusterOptions = markerClusterOptions(),
             popup = content$popup)

```

\*note: 976 sightings were removed from this interactive map as they did not provide the longitudes and latitudes.

```{r}
bigfoot_map
```

This map is nice because you can zoom in on each encounter and see *exactly* where each Bigfoot encounter occured. But because of the subgrouping of the pins, it is difficult to determine the geographic density of these sightings. It certainly *looks* like there may be some patterns. For example, in the Midwest, there appears to be a somewhat lower number of pins when you zoom out to look at the entire country, but it is difficult to determine for sure. We have the *stories*, but there is a limit to how much information we can fit into each pop-up, really with just space for the witness's written/spoken testimony.

To demonstrate the sighting density in each state, I now want to generate a map of North America where each of the 50 United States are shaded according to how many Bigfoot encounters have occured within it. If I were lucky, there would be no Bigfoot sightings in Alaska and I could simply use the United States map provided by the "maps" R package;

```{r}
bigfoot %>%
  filter(state == "Alaska" | state == "Hawaii") %>%
  group_by(state) %>%
  summarize(
    n_sightings = n()
  )
```

But alas. We would be missing out on a substantial 20 sightings by excluding Alaska.

So instead, I will need to use the R package USAboundaries to get the geometries for each of the 48 contiguous states and also Alaska and join this data with a dataset containing the number of sightings per state.

```{r}
states <- states(cb = TRUE)

library(USAboundaries)

states <- us_states() %>%
  filter(name != "Hawaii", name != "Puerto Rico") %>%
    st_crop(xmin = 130, xmax = -60,
            ymin = 0, ymax = 70)

sightings_by_state <- bigfoot %>%
  group_by(state) %>%
  summarize(
    n_sightings = n()
  )

sightings_joined <- full_join(states, sightings_by_state, join_by(name == state))

ggplot(data = sightings_joined) +
  geom_sf(aes(fill = n_sightings)) +
  scale_fill_viridis_c(direction = -1) +
  coord_sf(xlim = c(-170, -60), ylim = c(20, 73)) +
  labs(fill = "Number of Sightings") +
  theme_void()
```

Now this is where it really got interesting. Washington appears to have *by far* the most Bigfoot sightings, clocking in at over 600. What makes Washington different from other states? How dramatic is its lead? For my next visualization, I want to investigate these questions by generating a line graph that depicts the rise of bigfoot sightings over time in the top five states with the most sightings over all. We conveniently already have a date variable in the original dataset, so it's really just a matter of figuring out how to generate a line plot that shows the **cumulative** number of cases at each "point."

```{r}
sightings_by_state %>%
  slice_max(order_by = n_sightings, n = 5)
```

```{r}
sightings_over_time <- bigfoot %>%
  filter(!is.na(date)) %>%
  filter(state == "Washington"|
         state == "California" |
         state == "Ohio" |
         state == "Florida" |
         state == "Oregon") %>%
  mutate(
    date = ymd(date)
  ) %>%
  group_by(state, date) %>%
  summarize(
    n_sightings = n()
  ) %>%
  mutate(
    cumulative_sightings = cumsum(n_sightings)
    )


ggplot(sightings_over_time, aes(x = date, 
                                y = cumulative_sightings, 
                                color = state)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Bigfoot Sightings Over Time",
    x = "Date",
    y = "Total Sightings",
    color = "State"
  ) + 
  scale_color_manual(values = c("red","seagreen2","slateblue2","thistle2","yellow")) +
  scale_x_date(limit=c(as.Date("1950-01-01"),as.Date("2026-04-24"))) +
  labs(caption = "*7 sightings have been excluded as they occured before 01-01-1950. Axis scaled for ease of reading.") +
  theme_minimal() 
```

With this visualization, we can make a few minor observations. The first is that Washington has, more or less, always led the 50 states in Bigfoot sightings. The only time where Washington was not in the lead was in the 60s, when California briefly had more. Other interesting observations; ever since approximately 2005, sightings in both Oregon and California have slowed to an almost pleateaued rate. This is especially interesting because of these states' close proximity to Washington, which is still experiencing sightings at a higher rate than the other 4 states.

Why is this? There could be any number of reasons. Washington state is often regarded as having incredible wilderness. This, coupled with a large, outdoorsy population, may be one explanation. There are simply many people who enjoy being outside, which is, consequently, where Bigfoot also enjoys being.

It may also be a culture of superstition and belief in Bigfoot. A quick, brief google search on "Washington Bigfoot Culture" yielded results such as "Skagit-Squatch Musuem", "Metaline Falls Bigfoot Festival", and a Youtube channel called the "Olympic Project," based in WA, and "committed to documenting the existence of Sasquatch through science and education" (the channel has been inactive for 11 years but made headlines on squatchable.com in 2025 for their discovery of "bigfoot nests" located near a salmon-rich riverbed, believed to be one of their food sources.)

Regardless, with the large sample size to choose from, I believe Washington is a good candidate for my final visualization; a linear regression seeking to predict the "quality" of a sighting. We have been given 3 levels of quality; Class A (best), Class B (fair), and Class C(poor). For the sake of simplicity, however, I am going to code the quality as 1 (Class A, best), 2 (Class B, fair), and 3(Class C, worst).

Because the BFRO cites visibility as a major determinant of an encounter's quality, I am going to primarily use variables that I think impact a person's ability to accurately see a sasquatch through any kind of weather phenomenon, such as **visibility, precip_intensity,** and **precip_type.** Additionally, I want to use plotly to add some hovertext just giving some context to the points on plot, such as which **county** the encounter occurred in,

First, I have to tidy the data by filtering out just the cases in Washington, coding the **classification** variable as numeric, and coding NAs in the precip_type variable as "None," with the assumption (and hope) that most of them are coded as NA because there was no precipitation, not because the precipitation type was not recorded.

```{r}
bigfoot_quality <- bigfoot %>%
  filter(state == "Washington") %>%
  mutate(
    quality = case_when(
      classification == "Class A" ~ 1,
      classification ==  "Class B"~ 2,
      classification == "Class C" ~ 3
  ),
    precip_type = factor(case_when(
      precip_type == "rain" ~ "rain",
      precip_type == "snow" ~ "snow",
      is.na(precip_type) ~ "none"
    ))
  ) 

mod_quality <- lm(quality ~ visibility + precip_intensity + precip_type + precip_type:precip_intensity, data = bigfoot_quality)

pred_quality <- ggpredict(mod_quality, terms = c("visibility [fivenum]",
                              "precip_intensity [fivenum]",
                              "precip_type")) %>%
  as.data.frame(terms_to_colnames = TRUE) %>%
  as_tibble()

plot_quality <- ggplot(data = pred_quality, aes(x=visibility, y=predicted))+
  geom_line(aes(color = precip_intensity), linewidth = 1.5) +
  facet_wrap(~precip_type) +
  scale_color_brewer(palette = "Pastel1") +
  theme_dark() +
  geom_point(data = bigfoot_quality, aes(y=quality, 
                                         color = precip_type,
                                         Title = `title`,
                                         County = county,
                                         Classification = classification
                                         ), alpha = 0.5) +
  theme(base_size = 24)

ggplotly(plot_quality, tooltip = c("Title", "County", "Classification"))
```

So this graph is a little difficult to interpret, given the nature of the data, *however*, it is possible to make some interesting observations. For example, we see that there is only one Class C encounter recorded in Washington, and it was on a clear day with 10/10 visibility. As for the lines; unsurprisingly, clear days were closer to a classification of Class A than rainy days, across all visibilities and precipitation intensities. All around, it does look like as visibility increases, more encounters tend towards Class A. Interestingly, snowy days tended to be more credible than rainy days. My guess is that some of the evidence that the BFRO considered came in the form of traces left in the snow , like large footprints (as referenced by 2/7 of the points in the "snow" panel). There is also far less vegetation in the winter to obscure the big hairy form of a Bigfoot through the branches.

## **Conclusions**

Washington leads the U.S in the State with the most Bigfoot sightings, a pattern that has more or less existed since the first sighting in the early 1900s. Notably, Oregon and California have both appeared to be plateauing since roughly 2010, at an almost identical rate. Within Washington, the quality of Bigfoot sightings is significantly associated with the weather conditions at the time of the encounter, with snowy weather sightings and clear weather sightings having almost similar levels of credibility. It is worth noting, in connection with this, that when the weather is snowy, more intense precipitation leads to *better credibility* than sightings with *less intense* snowy conditions*.* This is different from rainy weather, where *more intense rain* is associated with *less credibility*. One potential explanation is that heavier snow is more likely to leave more defined footprints, which may contribute to the BFRO's assessment of credibility.

Given more time (and skill), there are certainly ways I would have liked to improve on this project. One of my classmates pointed out during my presentation that Washington may have so many sightings because Bigfoot encounters tend to occur in the mountains, of which there are many in WA. if I could have found a geographic map of the U.S., it might have been beneficial to try and join that elevation data into my dataset and see if higher elevations had more sightings. I also keep thinking that there must have been a better way to predict sighting credibility, but I just couldn't work it out in time. Overall, I think I overestimated the significance of the variables in this dataset, and perhaps would have tried to look for a different one on Bigfoot sightings.
